The Rockerverse: Packages and Applications for Containerisation with R

### 2020-02-04 manuscript.Rmd
fix Dirk's email, add snipped for email list
proofreading changes

title: "The Rockerverse: Packages and Applications for Containerization with R"

@@

    email: dirkd@eddelbuettel.com

@@

  The Rocker Project provides widely-used Docker images for R across different application scenarios.
  This articles surveys downstream projects building upon Rocker and presents the current state of R packages for managing Docker images and controlling containers.

@@

  We identified common themes across this diversity: reproducible environments, scalability and efficiency, and portability across clouds.
# blank footnote? https://tex.stackexchange.com/questions/170511/footnotes-without-numbering

@@

The R community keeps growing.
This can be seen in the number of new packages on CRAN, which keeps on growing exponentially \citep{cran:2019}, but also in the numbers of conferences, open educational resources, meetups, unconferences, and companies taking up, as exemplified by the useR! conference series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}}, the global growth of the R and R-Ladies user groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}}, or the foundation and impact of the R&nbsp;Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
All this cements the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
Coinciding with the rise of R was the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distribution and deployment of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next Section for more details).
The considerable uptake and continued evolution of the Rocker&nbsp;Project has lead to numerous projects extending or building upon Rocker images, ranging from reproducible research\footnote{"Reproducible" in the sense of the \emph{Claerbout/Donoho/Peng} terminology \citep{barba_terminologies_2018}.} to production deployments.
This article presents this _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
We also introduce related activities connecting the R language and environment with other containerisation solutions.
The main contributions is a coherent picture of the current lay of the land of using containers in, with, and for R.

The article continues with a brief introduction of containerization basics and the Rocker&nbsp;Project, followed by use cases and applications, starting with the R packages specificly for interacting with Docker, second-level packages using containers indirectly or only for specific features, up to complex use cases leveraging containers.
We conclude with a reflection on the landscape of packages and applications and point out future directions of development.

# Containerization and Rocker
\label{containerisation}
\label{rocker}

Docker, an application and service provide by the eponymous company, has in just a few short years risen to prominence for development, testing, deployment and distribution of computer software \citep[cf.][]{datadog_8_2018,munoz_history_2019}.
While there are related approaches such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synomymous with "containerization"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operationg systems.
In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualization.
By allowing (one or possibly) multiple applications or services to run concurrently on one host machine without any fear of interference between them, an important scalability opportunity is being provided.
But Docker improved this compartmentalization by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualization.
This containerization is also called operating-system-level virtualization \citep{wikipedia_contributors_os-level_2020}.
Typically a container runs one process, whereas virtualization may run whole operating systems at a larger footprint.
This makes for more efficient use of system resources \citep{felter_updated_2015} and allowed another order of magnitude in terms of scalability of deployment \citep[cf.][]{datadog_8_2018}.
While Docker makes use of Linux kernel features, its importance was large enough so that some required aspects of running Docker have been added to other operating systems to support Docker there more efficiently too \citep{microsoft_linux_2019}.
The success even lead to standardisation and industry collaboration \citep{oci_open_2019}.

The key accomplishment of Docker as an "application" is to make a "bundled" aggregation of software (the so-called "image") available to any system equipped to run Docker, without requiring much else from the host besides the actual Docker application installation.
This is a rather attractive proposition and Docker's very easy to use user interface has lead to widespread adoption and use of Docker in a variety of domains, e.g., cloud computing infrastructure \citep[e.g.,][]{Bernstein2014}, data science \citep[e.g.,][]{boettiger_introduction_2015}, and edge computing \citep[e.g.,][]{alam_orchestration_2018}.
It provided to be a natural match for "cloud deployment" which runs, or at least appears to run, "seamlessly" without much explicit reference to the underlying machine, architecture or operating system: containers are portable and can be deployed with very little in terms of dependencies on the host system---only the container runtime is required.
Images are normally built from plain text documents called `Dockerfile`.
A `Dockerfile` has a specific set of instructions to create and document a well-defined environment, i.e., install specific software and expose specific ports.

For statistical computing and analysis centered around R, the **Rocker&nbsp;Project** has provided a variety of Docker containers since its start in 2014 \citep{RJ-2017-065}.
The Rocker&nbsp;Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
These containers form _image stacks_, building on top of each other for better maintainability (i.e., smaller `Dockerfile`s), composability, and to reduce build time.
Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
The Rocker&nbsp;Project's impact and importance was acknowledged by the Chan Zuckerberg Initiative's _Essential Open Source Software for Science_, who provide findung for the projects's sustainable maintenance, community growth, and targeting new hardware platforms including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

Docker is not the only containerisation software.
An alternative stemming from the domain of high-perfomance computing is **Singularity** \citep{kurtzer_singularity_2017}.
Singularity can run Docker images, and in the case of Rocker works out of the box if the main process is R, e.g., in `rocker/r-base`, but does not succeed in running images where there is an init script, e.g., in containers that by default run RStudio&nbsp;Server.
In the latter case, a `Singularity` file, a recipe akin to a `Dockerfile`, needs to be used.
To date, no comparable image stack to Rocker exists on [Singularity Hub](https://singularity-hub.org/).
A further tool for running containers is [**podman**](https://github.com/containers/libpod), which also can build `Dockerfile`s and run Docker images.
Proof of concepts for using podman to build and run Rocker containers exist\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}}.
Yet the prevelance of Docker, especially in the broader user community beyond experts or niche systems, and the vast amount of blog posts and courses for Docker, currently caps specific development efforts for both Singularity and podman in the R community.
This might quickly change when usability and spread increase, or security features such as rootless/unprivileged containers, which both these tools support out of the box, become more sought after.

@@

Interfacing with the Docker daemon is typically done through the [Docker Command Line Interface](https://docs.docker.com/engine/reference/commandline/cli/) (Docker CLI).
However, moving back and forth between an R console and the command line can create friction in workflows and reduce reproducibility.
A number of first-order R packages provide a interface to the Docker CLI, allowing to automate interaction with the Docker CLI from an R console.

Each of these packages has particular advantages as they provide function wrappers for interacting with the Docker CLI at different stages of a container's life cycle.
Examples of such interactions are installing the Docker software, creating `Dockerfile`s (\CRANpkg{dockerfiler}, \pkg{containerit}), building images and launching a containers (\CRANpkg{stevedore}, \pkg{docker}) on a local machine or on the cloud.
As such, the choice of which package is most useful depends on the use-case at hand, but also the users level of expertise.

@@

# some functionalities might be missing. please add them if you find that any are missing!
package_functionalities <- tribble(
  ~package, ~id,
  "stevedore", c("api", "execute", "cloud", "images", "interact"), # build?
  "dockyard", c("dockerfile", "build", "execute"), # cloud?
  "dockermachine", c("install", "execute", "cloud", "interact"),
  "AzureContainers", c("build", "execute", "cloud"),
  "babelwhale", c("execute", "interact", "othercontainers"),  
  "harbor", c("execute", "cloud", "images")
)

@@

**\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor) is not actively maintained anymore, but should be honorably mentioned as the first R package for managing Docker images and containers.
It uses the \CRANpkg{sys} package to run system commands against the Docker CLI, both locally and through an SSH connection, and has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
The output of container executions are converted to appropriate R types.
The Docker CLI's basic functionality, while evolving quickly and with small concern for avoiding breaking changes, is unchanged for a long time so a core function such as `harbor::docker_run(image = "hello-world")` still works depsite the stopped development.
```{r harbor, eval=FALSE, include=FALSE}
remotes::install_github("wch/harbor")
harbor::docker_run(image = "hello-world")
```

**\CRANpkg{stevedore}** ([https://cran.r-project.org/package=stevedore](https://cran.r-project.org/package=stevedore)) is currently the most powerful Docker client in R.
It interfaces with the Docker daemon over the Docker HTTP API\footnote{\href{https://docs.docker.com/engine/api/latest/}{https://docs.docker.com/engine/api/latest/}} via a Unix socket on Linux or MacOS, over a named pipe on Windows, or over an HTTP/TCP connection.
The package is the only one not using system calls to the `docker` CLI for managing images and containers and easily exposes connections to remote Docker daemons, which has to be configured on the Docker level otherwise.
Using the API gives access to more information and is system independent and likely more reliable than parsing command line output.
\CRANpkg{stevedore}'s own interface is automatically generated based on the OpenAPI specification of the Docker daemon, but still similar to the Docker CLI.
The interface is similar to R6 objects, in that a `stevedore_object` has a number of functions attached to it that can be called, and multiple specific versions of the Docker API can be supported thanks to the automatic generation\footnote{See \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.

@@

**\CRANpkg{babelwhale}** allows executing and interacting with containers, which can use either Docker or Singularity as a backend \citep{cannoodt_babelwhale_2019}.
The package provides a unified interface to interact with Docker and Singularity containers.
Users can, for example, execute a command inside a container, mount a volume or copy a file.

**\pkg{dockyard}** ([https://github.com/thebioengineer/dockyard](https://github.com/thebioengineer/dockyard)) has the goal of lowering barrier to creating `Dockerfile`s, building Docker images, and deploying Docker containers.
The package follows the increasingly used piping paradigm of the `tidyverse` style of programming for chaining R functions representing the instructions in a `Dockerfile`.
An existing `Dockerfile` can be used as a template.
\pkg{dockyard} also includes wrappers for common steps, such as installing an R package or copying files, and build-in functions for building an image running a container, to make using Docker even more approachable to R users with a single API.

**\pkg{dockermachine}** ([https://github.com/cboettig/dockermachine](https://github.com/cboettig/dockermachine)) is an R package to provide a convenient interface to [Docker&nbsp;Machine](https://docs.docker.com/machine/overview/) from R.
The CLI tool `docker-machine` allows users to create and manage virtual host on local computers, local data centers, or at cloud providers.
A local Docker installation can be configured to transparntly forward all commands issued on the local Docker CLI to a selected (remote) virtual host.
Docker&nbsp;Machine was especially crucial for local use in early days of Docker, when no native support was available for Mac or Windows computers, but remains relevant for provisioning on remote systems.
The package has not received any updates for two years, but is functional with a current version of `docker-machine` (`0.16.2`).
It potentially lowers the barriers for R users to run containers on various hosts, if using the Docker&nbsp;Machine CLI directly is perceived as a barrier, or enables scripted workflows with remote processing.
```{r dockermachine, eval=FALSE, include=FALSE}
library("dockermachine")
dockermachine::machine_create(driver = "amazonec2")
# machine started: https://console.aws.amazon.com/ec2/v2/home?region=us-east-1#Instances:sort=instanceId
# get config on command line: $ docker-machine env machine
# activate: $ $(docker-machine env test) does not work, because of "
# run container: docker run -it -p 80:80 nginx
# allow inbound HTTP traffic
# http://ec2-3-85-85-61.compute-1.amazonaws.com/ shows "Welcome to nginx"!
# Nice!
# > dockermachine::machine_rm()
# Instance shut down.
```

@@

**Bioconductor** ([https://bioconductor.org/](https://bioconductor.org/)) is an open source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
The project consists of 1823 R software packages as of October 30th 2019, as well as packages containing annotation or experiment data.
_Bioconductor_ has a semi-annual release cycle, each release is associated with a particular version of R, and Docker images are provided for current and past versions of _Bioconductor_ for convenience and reproducibility.
All images, included  are described on the _Bioconductor_ web site (see [https://bioconductor.org/help/docker/](https://bioconductor.org/help/docker/)), created with `Dockerfile`s maintained on GitHub, and distributed through Docker&nbsp;Hub\footnote{See \href{https://github.com/Bioconductor/bioconductor_docker}{https://github.com/Bioconductor/bioconductor\_docker} and \href{https://hub.docker.com/u/bioconductor}{https://hub.docker.com/u/bioconductor} respectively.}.
_Bioconductor_'s 'base' Docker images are built on top of the `rocker/rstudio` image.
_Bioconductor_ installs packages based on the R version in combination with the Bioconductor version, and therefore uses Bioconductor version tagging `devel` and `RELEASE_X_Y`, e.g., `RELEASE_3_10`.
Past and current combinations of R and _Bioconductor_ will therefore be accessible via a specific image tags.
The _Bioconductor_ `Dockerfile` selects the desired version of R from Rocker, adds required system dependencies, and uses the \CRANpkg{BiocManager} package for installing appropriate versions of _Bioconductor_ packages.
A strength of this approach is that the responsibility for complex software configuration (including customized development) is shifted from the user to the experienced _Bioconductor_ core team.
A recent audit of the _Bioconductor_  image stack `Dockerfile` led to the deprecation of several community-maintained images, because the numerous specific images became too hard to understand, complex to maintain, and cumbersome to extent.
As part of the simplification, a recent innovation is to produce a `bioconductor_docker:devel` image to emulate the _Bioconductor_ nightly Linux build machine as closely as possible.
This image contains the build system environment variables and the _system dependencies_ needed to install and check almost all (1813 out of 1823) _Bioconductor_ software packages and saves users and package developers from managing these themselves.
Furthermore the image is configured so that `.libPaths()` has `/usr/local/lib/R/host-site-library` as the first location.
Users mounting a location on the host file system to this location can persist installed packages across Docker sessions or image updates.
Many R users pursue flexible work flows tailored to particular analysis needs, rather than standardized work flows.
The new `bioconductor_docker` image is well-suited for this pattern, while `bioconductor_docker:devel` provides developers with a test environment close to _Bioconductor_'s build system.

\label{datascience}
**Data Science** is a widely discussed topic among all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
The discussions shed a light on the tools and craftspersonship behind the analysis of data with computational methods.
The practice of Data Science often involves a combination tools and software stacks and requires a cross-cutting skillset.
This complexity and an inherent concern for openness and reproducibility in the Data Science community lead to Docker being used widely.
<!--A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).-->
The remaineder of this section presents exemplary Docker images and image stacks featuring R intended for Data Science.
The [_Jupyter Docker Stacks_](https://github.com/jupyter/docker-stacks/) project are a set of ready-to-run Docker images containing Jupyter applications and interactive computing tools \citep{project_jupyter_jupyter_2018}.
The `jupyter/r-notebook` image includes R and "popular packages", and naturally also the IRKernel ([https://irkernel.github.io/](https://irkernel.github.io/)), an R kernel for Jupyter so that Jupyter Notebooks can contain R code cells.

@@

R is installed using the Conda package manager, which can manage environments for various programming languages, pinning both the R version and the versions of R packages\footnote{See \texttt{jupyter/datascience-notebook}'s \texttt{Dockerfile} at \href{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}.}.
_Kaggle_ provides the [`gcr.io/kaggle-images/rstats`](https://hub.docker.com/r/kaggle/rstats) image (previously `kaggle/rstats`) and [corresponding `Dockerfile`](https://github.com/Kaggle/docker-rstats) for usage in their Machine Learning competitions and easy access to the associated datasets.
It includes machine learning libraries such as Tensorflow and Keras (see also image `rocker/ml` in Section&nbsp;\nameref{rocker-gpu}), and also configures the \CRANpkg{reticulate} package.

>>>>

title: "The Rockerverse: Packages and Applications for Containerisation with R"

@@

    email: dirk@eddelbuettel.com

@@

  The Rocker Project provides widely used Docker images for R across different application scenarios.
  This articles surveys downstream projects that build upon Rocker images and presents the current state of R packages for managing Docker images and controlling containers.

@@

  Across the diverse ways to use containers, we identified common themes: reproducible environments, scalability and efficiency, and portability across clouds.

@@

The R community continues to grow.
This can be seen in the number of new packages on CRAN, which is still on growing exponentially \citep{cran:2019}, but also in the numbers of conferences, open educational resources, meetups, unconferences, and companies that are adopting R, as exemplified by the useR! conference series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}}, the global growth of the R and R-Ladies user groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}}, or the foundation and impact of the R&nbsp;Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
These trends cement the role of R as the _lingua&nbsp;franca_ of statistics, data visualisation, and computational research.
Also though, coinciding with the rise of R has been the advent of [Docker](https://en.wikipedia.org/wiki/Docker_(software)) as a general tool for distributing and deploying of server applications---in fact, Docker can be called the _lingua&nbsp;franca_ of describing computing environments and packaging software.
Combining both these topics, the _Rocker&nbsp;Project_ ([https://www.rocker-project.org/](https://www.rocker-project.org/)) provides images with R (see the next section for more details).
The considerable uptake and continued evolution of the Rocker&nbsp;Project has led to numerous projects that extend or build upon Rocker images, ranging from reproducible\footnote{"Reproducible" in the sense of the \emph{Claerbout/Donoho/Peng} terminology \citep{barba_terminologies_2018}.} research to production deployments.
As such, this article presents this _Rockerverse_ of projects across all development stages: early demonstrations, working prototypes, and mature products.
We also introduce related activities that connect the R language and environment with other containerisation solutions.
Our main contributions is a coherent picture of the current status of using containers in, with, and for R.

The article continues with a brief introduction of containerisation basics and the Rocker&nbsp;Project, followed by use cases and applications, starting with the R packages specificly for interacting with Docker, next the second-level packages that use containers indirectly or only for specific features, and finally some complex use cases that leverage containers.
We conclude with by reflecting on the landscape of packages and applications and point out future directions of development.

# Containerisation and Rocker
\label{containerisation}
\label{rocker}

Docker, an application and service provided by the eponymous company, has, in just a few short years, risen to prominence for developing, testing, deploying and distributing computer software \citep[cf.][]{datadog_8_2018,munoz_history_2019}.
While related approaches exist, such as LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}} or Singularity \citep{kurtzer_singularity_2017}, Docker has become synomymous with "containerisation"---the method of taking software artefacts and bundling them in such a way that use becomes standardized and portable across operationg systems.
In doing so, Docker had recognised and validated the importance of one very important thread that had been emerging, namely virtualisation.
By allowing (one or possibly) multiple applications or services to run concurrently on one host machine without any fear of interference between them, Docker provides an important scalability opportunity.
Beyond this though, Docker has improved this compartmentalisation by accessing the host system---generally Linux---through a much thinner and smaller shim than a full operating system emulation or virtualisation.
This containerisation, also called operating-system-level virtualisation \citep{wikipedia_contributors_os-level_2020}, 
makes efficient use of operating system resources \citep{felter_updated_2015} and allows another order of magnitude in terms of scalability of deployment \citep[cf.][]{datadog_8_2018} because, while a container typically runs one process, virtualisation may run whole operating systems at a larger footprint.
This makes for more efficient use of system resources \citep{felter_updated_2015} and allowed another order of magnitude in terms of scalability of deployment \citep[cf.][]{datadog_8_2018}.
While Docker makes use of Linux kernel features, it has become important enough that some required aspects of running Docker have been added to other operating systems so that those systems can more efficiently support Docker \citep{microsoft_linux_2019}.
The success of Docker has even led to standardisation and industry collaboration \citep{oci_open_2019}.

The key accomplishment of Docker as an "application" is to make a "bundled" aggregation of software, the so-called "image", available to any system equipped to run Docker, without requiring much else from the host besides the actual Docker application installation.
This is a rather attractive proposition, and Docker's very easy to operate user interface has led to widespread adoption and use of Docker in a variety of domains, e.g., cloud computing infrastructure \citep[e.g.,][]{Bernstein2014}, data science \citep[e.g.,][]{boettiger_introduction_2015}, and edge computing \citep[e.g.,][]{alam_orchestration_2018}.
It has also proven to be a natural match for "cloud deployment" which runs, or at least appears to run, "seamlessly" without much explicit reference to the underlying machine, architecture or operating system:
Containers are portable and can be deployed with very little dependencies on the host system---only the container runtime is required.
These Docker images are normally built from plain text documents called `Dockerfile`s; a `Dockerfile` has a specific set of instructions to create and document a well-defined environment, i.e., install specific software and expose specific ports.

For statistical computing and analysis centered around R, the **Rocker&nbsp;Project** has provided a variety of Docker containers since it began in 2014 \citep{RJ-2017-065}.
The Rocker&nbsp;Project provides several lines of containers spanning to from building blocks with `R-release` or `R-devel`, via containers with [RStudio&nbsp;Server](https://rstudio.com/products/rstudio/) and [Shiny&nbsp;Server](https://rstudio.com/products/shiny/shiny-server/), to domain-specific containers such as [`rocker/geospatial`](https://github.com/rocker-org/geospatial) \citep{rocker_geospatial_2019}.
These containers form _image stacks_, building on top of each other for easier maintainability (i.e., smaller `Dockerfile`s), better composability, and to reduce build time.
Also of note is a series of "versioned" containers which match the R release they contain with the _then-current_ set of packages via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}.
The Rocker&nbsp;Project's impact and importance was acknowledged by the Chan Zuckerberg Initiative's _Essential Open Source Software for Science_, which provides funding for the projects's sustainable maintenance, community growth, and targeting new hardware platforms including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

Docker is not the only containerisation software.
An alternative stemming from the domain of high-perfomance computing is **Singularity** \citep{kurtzer_singularity_2017}.
Singularity can run Docker images.
Rocker images work out of the box if the main process is R, e.g., in `rocker/r-base`, but Singularity does not succeed in running images where there is an init script, e.g., in containers that by default run RStudio&nbsp;Server.
In the latter case, a `Singularity` file, a recipe akin to a `Dockerfile`, needs to be used to make necessary adjustments.
To date, no comparable image stack to Rocker exists on [Singularity Hub](https://singularity-hub.org/).
A further tool for running containers is [**podman**](https://github.com/containers/libpod), which can also build `Dockerfile`s and run Docker images.
Proof of concepts exists for using podman to build and run Rocker containers\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}}, but the prevelance of Docker, especially in the broader user community beyond experts or niche systems and the vast amount of blog posts and courses for Docker currently cap specific development efforts for both Singularity and podman in the R community.
This might quickly change if the usability and spread of Singularity or podman increase, or if security features such as rootless/unprivileged containers, which both these tools support out of the box, become more sought after.

@@

Interfacing with the Docker daemon is typically done through the [Docker Command Line Interface](https://docs.docker.com/engine/reference/commandline/cli/) (Docker CLI).
However, moving back and forth between an R console and the command line can create friction in workflows and reduce reproducibility.
A number of first-order R packages provide an interface to the Docker CLI, allowing for the interaction with the Docker CLI from an R console, e.g., to be automated.

Each of these packages has particular advantages as they provide function wrappers for interacting with the Docker CLI at different stages of a container's life cycle.
Examples of such interactions are installing the Docker software, creating `Dockerfile`s (\CRANpkg{dockerfiler}, \pkg{containerit}), building images, and launching a containers (\CRANpkg{stevedore}, \pkg{docker}) on a local machine or on the cloud.
As such, the choice of which package is most useful depends on the use case at hand as well as on the user's level of expertise.

@@

# some functionalities might be missing. please add them if you find that any are missing!
package_functionalities <- tribble(
  ~package, ~id,  
  "harbor", c("execute", "cloud", "images"),
  "stevedore", c("api", "execute", "cloud", "images", "interact"), # build?
  "AzureContainers", c("build", "execute", "cloud"),
  "babelwhale", c("execute", "interact", "othercontainers"),
  "dockyard", c("dockerfile", "build", "execute"), # cloud?
  "dockermachine", c("install", "execute", "cloud", "interact")
)

@@

**\pkg{harbor}** ([https://github.com/wch/harbor](https://github.com/wch/harbor) is no longer actively maintained, but it should be honorably mentioned as the first R package for managing Docker images and containers.
It uses the \CRANpkg{sys} package to run system commands against the Docker CLI, both locally and through an SSH connection, and it has convenience functions, e.g., for listing and removing containers/images and for accessing logs.
The output of container executions are converted to appropriate R types.
The Docker CLI's basic functionality, although it evolves quickly and with little concern for avoiding breaking changes, has remained unchanged in core functions, meaning that a core function such as `harbor::docker_run(image = "hello-world")` still works depsite its stopped development.
```{r harbor, eval=FALSE, include=FALSE}
remotes::install_github("wch/harbor")
harbor::docker_run(image = "hello-world")
```

**\CRANpkg{stevedore}** ([https://cran.r-project.org/package=stevedore](https://cran.r-project.org/package=stevedore)) is currently the most powerful Docker client in R.
It interfaces with the Docker daemon over the Docker HTTP API\footnote{\href{https://docs.docker.com/engine/api/latest/}{https://docs.docker.com/engine/api/latest/}} via a Unix socket on Linux or MacOS, over a named pipe on Windows, or over an HTTP/TCP connection.
The package is the only one not using system calls to the `docker` CLI tool for managing images and containers.
The package thereby enables connections to remote Docker instances without direct configuration of the local Docker daemon.
Furthermore using the API gives access to information in a structured way, is system independent, and is likely more reliable than parsing command line output.
\CRANpkg{stevedore}'s own interface is automatically generated based on the OpenAPI specification of the Docker daemon, but it is still similar to the Docker CLI.
The interface is similar to R6 objects, in that a `stevedore_object` has a number of functions attached to it that can be called, and multiple specific versions of the Docker API can be supported thanks to the automatic generation\footnote{See \href{https://github.com/richfitz/stevedore/blob/master/development.md}{https://github.com/richfitz/stevedore/blob/master/development.md}.}.

@@

**\CRANpkg{babelwhale}** provides a unified interface to interact with Docker and Singularity containers \citep{cannoodt_babelwhale_2019}.
Users can, for example, execute a command inside a container, mount a volume, or copy a file with the same R commands for both container runtimes.

**\pkg{dockyard}** ([https://github.com/thebioengineer/dockyard](https://github.com/thebioengineer/dockyard)) has the goal of lowering the barrier to creating `Dockerfile`s, building Docker images, and deploying Docker containers.
The package follows the increasingly used piping paradigm of the `tidyverse` style of programming for chaining R functions representing the instructions in a `Dockerfile`.
An existing `Dockerfile` can be used as a template.
\pkg{dockyard} also includes wrappers for common steps, such as installing an R package or copying files, and provides built-in functions for building an image and running a container, which make Docker more approachable within a single R-based user interface.

**\pkg{dockermachine}** ([https://github.com/cboettig/dockermachine](https://github.com/cboettig/dockermachine)) is an R package to provide a convenient interface to [Docker&nbsp;Machine](https://docs.docker.com/machine/overview/) from R.
The CLI tool `docker-machine` allows users to create and manage a virtual host on local computers, local data centers, or at cloud providers.
A local Docker installation can be configured to transparntly forward all commands issued on the local Docker CLI to a selected (remote) virtual host.
Docker&nbsp;Machine was especially crucial for local use in the early days of Docker, when no native support was available for Mac or Windows computers, but it remains relevant for provisioning on remote systems.
The package has not received any updates for two years, but it is functional with a current version of `docker-machine` (`0.16.2`).
It potentially lowers the barriers for R users to run containers on various hosts if they perceive that using the Docker&nbsp;Machine CLI directly as a barrier and it enables scripted workflows with remote processing.
```{r dockermachine, eval=FALSE, include=FALSE}
library("dockermachine")
dockermachine::machine_create(driver = "amazonec2")
# machine started: https://console.aws.amazon.com/ec2/v2/home?region=us-east-1#Instances:sort=instanceId
# get config on command line: $ docker-machine env machine
# activate: $ $(docker-machine env test) does not work, because of "
# run container: docker run -it -p 80:80 nginx
# allow inbound HTTP traffic
# http://ec2-3-85-85-61.compute-1.amazonaws.com/ shows "Welcome to nginx"!
# Nice!
# > dockermachine::machine_rm()
# Instance shut down.
```

@@

**Bioconductor** ([https://bioconductor.org/](https://bioconductor.org/)) is an open source, open development project for the analysis and comprehension of genomic data \citep{gentleman_bioconductor_2004}.
As of October 30th 2019, the project consists of 1823 R software packages, as well as packages containing annotation or experiment data.
_Bioconductor_ has a semi-annual release cycle, where each release is associated with a particular version of R, and Docker images are provided for current and past versions of _Bioconductor_ for convenience and reproducibility.
All images, which are described on the _Bioconductor_ web site (see [https://bioconductor.org/help/docker/](https://bioconductor.org/help/docker/)), are created with `Dockerfile`s maintained on GitHub and distributed through Docker&nbsp;Hub\footnote{See \href{https://github.com/Bioconductor/bioconductor_docker}{https://github.com/Bioconductor/bioconductor\_docker} and \href{https://hub.docker.com/u/bioconductor}{https://hub.docker.com/u/bioconductor} respectively.}.
_Bioconductor_'s "base" Docker images are built on top of the `rocker/rstudio` image.
_Bioconductor_ installs packages based on the R version in combination with the Bioconductor version and, therefore, uses Bioconductor version tagging `devel` and `RELEASE_X_Y`, e.g., `RELEASE_3_10`.
Past and current combinations of R and _Bioconductor_ will therefore be accessible via a specific image tags.

The _Bioconductor_ `Dockerfile` selects the desired version of R from Rocker, adds required system dependencies, and uses the \CRANpkg{BiocManager} package for installing appropriate versions of _Bioconductor_ packages.
A strength of this approach is that the responsibility for complex software configuration (including customized development) is shifted from the user to the experienced _Bioconductor_ core team.
However, a recent audit of the _Bioconductor_  image stack `Dockerfile` led to the deprecation of several community-maintained images, because the numerous specific images became too hard to understand, complex to maintain, and cumbersome to customise.
As part of the simplification, a recent innovation is to produce a `bioconductor_docker:devel` image to emulate the _Bioconductor_ nightly Linux build machine as closely as possible.
This image contains the build system environment variables and the _system dependencies_ needed to install and check almost all (1813 out of 1823) _Bioconductor_ software packages, and it saves users and package developers from managing these themselves.
Furthermore, the image is configured so that `.libPaths()` has `/usr/local/lib/R/host-site-library` as the first location.
Users mounting a location on the host file system to this location can persistently manage installed packages across Docker containers or image updates.
Many R users pursue flexible work flows tailored to particular analysis needs rather than standardized work flows.
The new `bioconductor_docker` image is well suited for this preference, while `bioconductor_docker:devel` provides developers with a test environment close to _Bioconductor_'s build system.

\label{datascience}
**Data science** is a widely discussed topic in all academic disciplines \citep[e.g.,][]{donoho_50_2017}.
These discussions have shed a light on the tools and craftspersonship behind the analysis of data with computational methods.
The practice of data science often involves combining tools and software stacks and requires a cross-cutting skillset.
This complexity and an inherent concern for openness and reproducibility in the data science community led to Docker being used widely.
<!--A general description of what to consider when creating a Docker image used for data science projects in R is provided on [environments.rstudio.com](https://environments.rstudio.com/docker).-->
The remaineder of this section presents exemple Docker images and image stacks featuring R intended for data science.

- The [_Jupyter Docker Stacks_](https://github.com/jupyter/docker-stacks/) project is a set of ready-to-run Docker images containing Jupyter applications and interactive computing tools \citep{project_jupyter_jupyter_2018}.
The `jupyter/r-notebook` image includes R and "popular packages", and naturally also the IRKernel ([https://irkernel.github.io/](https://irkernel.github.io/)), an R kernel for Jupyter, so that Jupyter Notebooks can contain R code cells.

@@

R is installed using the Conda package manager\footnote{\href{https://conda.io/}{https://conda.io/}}, which can manage environments for various programming languages, pinning both the R version and the versions of R packages\footnote{See \texttt{jupyter/datascience-notebook}'s \texttt{Dockerfile} at \href{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}{https://github.com/jupyter/docker-stacks/blob/master/datascience-notebook/Dockerfile\#L47}.}.
- _Kaggle_ provides the [`gcr.io/kaggle-images/rstats`](https://hub.docker.com/r/kaggle/rstats) image (previously `kaggle/rstats`) and [corresponding `Dockerfile`](https://github.com/Kaggle/docker-rstats) for usage in their Machine Learning competitions and easy access to the associated datasets.
It includes machine learning libraries such as Tensorflow and Keras (see also image `rocker/ml` in Section&nbsp;\nameref{rocker-gpu}), and it also configures the \CRANpkg{reticulate} package.
